---
title: "Weather observation data"
author: "Joona Lehtomäki"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
vignette: >
  %\VignetteIndexEntry{Weather observation data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

`fmi2` is not yet in CRAN, so you'll need to install it directly from GitHub. 
While you're at it, make sure you also install all the packages below as we'll
be using them in this tutorial.

```{r installs, eval=FALSE}
install.packages(c("DT", "leaflet", "remotes", "tidyverse")

remotes::install_github("ropengov/fmi2")
remotes::install_github("ropensci/skimr")
```

```{r load-libraries}
library(DT)
library(fmi2)
library(leaflet)
library(skimr)
```

# Getting started

You can retrieve weather stating observation data with various temporal 
resolution using `fmi2`. We'll start with `obs_weather_daily()` which returns
daily average observation data from a given location. First thing you need to 
know is of course which location exactly you want to get the data from. The FMI
API provides multiple different ways of defining the spatial query area:

+ **Bounding box** given by coordindates and defining an area
+ **Place name** for which to provide data.
+ **FMISID** numeric FMI observation station identifier
+ **GEOID** numeric geoid of the location
+ **WMO** code of the location

We'll start off by using the **FMISID** identifies which is given to each [FMI
observation stations](https://en.ilmatieteenlaitos.fi/observation-stations). The
online table is also available in `fmi2` using the function `fmi_stations()`:

```{r show-stations}
station_data <- fmi2::fmi_stations() 

station_data %>% 
  DT::datatable()

```

We're going to pick "Hanko Tulliniemi" as an example here and use its FMISID
(100946) to retrieve the data. As you can see from the table above, it also 
provides the latlon (geographical) coordinates for the observation station.
Before we get the actual data, let's visualize Hanko region.

```{r plot-hanko-map}
# Get data for Tulliniemi only
tulliniemi_station <- station_data %>% 
  dplyr::filter(FMISID == 100946)

# Plot on a map using leaflet
leaflet::leaflet(station_data) %>% 
  leaflet::setView(lng = tulliniemi_station$Lon, 
                   lat = tulliniemi_station$Lat, 
                   zoom = 11) %>% 
  leaflet::addTiles() %>%  
  leaflet::addMarkers(~Lon, ~Lat, popup = ~Name, label = ~as.character(FMISID))
```


```{r getting-data}


# Use Hanko Tulliniemi weather station
hanko_id <- 100946

obs_dat <- obs_weather_daily(starttime = "2019-01-01",
                             endtime = "2019-06-01",
                             fmisid = hanko_id)

```


