---
title: "Weather observation data"
author: "Joona Lehtomäki"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
vignette: >
  %\VignetteIndexEntry{Weather observation data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

`fmi2` is not yet in CRAN, so you'll need to install it directly from GitHub. 
While you're at it, make sure you also install all the packages below as we'll
be using them in this tutorial.

```{r installs, eval=FALSE}
install.packages(c("DT", "leaflet", "remotes", "sf", "tidyverse")

remotes::install_github("ropengov/fmi2")
remotes::install_github("ropensci/skimr")
```

```{r load-libraries}
library(DT)
library(fmi2)
library(leaflet)
library(sf)
library(skimr)
```

# Getting started

You can retrieve weather stating observation data with various temporal 
resolution using `fmi2`. First thing you need to know is of course which 
location exactly you want to get the data from. The FMI API provides multiple 
different ways of defining the spatial query area:

+ **Bounding box** given by coordindates and defining an area
+ **Place name** for which to provide data.
+ **FMISID** numeric FMI observation station identifier
+ **GEOID** numeric geoid of the location
+ **WMO** code of the location

We'll start off by using the **FMISID** identifies which is given to each [FMI
observation stations](https://en.ilmatieteenlaitos.fi/observation-stations). The
online table is also available in `fmi2` using the function `fmi_stations()`:

```{r show-stations}
station_data <- fmi2::fmi_stations() 

station_data %>% 
  DT::datatable()

```

We're going to pick "Hanko Tulliniemi" as an example here and use its FMISID
(100946) to retrieve the data. As you can see from the table above, it also 
provides the latlon (geographical) coordinates for the observation station.
Before we get the actual data, let's visualize Hanko region.

```{r plot-hanko-map}
# Get data for Tulliniemi only
tulliniemi_station <- station_data %>% 
  dplyr::filter(FMISID == 100946)

# Plot on a map using leaflet
leaflet::leaflet(station_data) %>% 
  leaflet::setView(lng = tulliniemi_station$Lon, 
                   lat = tulliniemi_station$Lat, 
                   zoom = 11) %>% 
  leaflet::addTiles() %>%  
  leaflet::addMarkers(~Lon, ~Lat, popup = ~Name, label = ~as.character(FMISID))
```


## Getting daily weather observation data

Now that we know how the address a specific observation station, we can proceed
to getting the actual data. `fmi2` providers several functions to retrieving 
data with different variables and temporal resolution. We'll start with 
`obs_weather_daily()` which returns daily average observation data from a given 
location. Let's get the daily weather observation data for the first 6 monhts
of 2019:

```{r getting-data}
# Use Hanko Tulliniemi weather station FMISID
obs_data <- obs_weather_daily(starttime = "2019-01-01",
                              endtime = "2019-06-30",
                              fmisid = 100946)
```

In total, the function returned `r nrow(obs_data)` observations. You can also
note the following:

```{r sf-class}
class(obs_data)
```

which means that the data returned by `obs_weather_daily()` is a spatial `sf`
object with the `geometry` column storing the geographical information of the
weather station. We'll come back to this later. Now we are interested in what 
*kind* of data did we actually get? Let's find out:

```{r}
unique(obs_data$variable)
```

So there are six variables with their corresponding values. `fmi2` provides a
helper function `describe_variables()` that can be useful in finding out more
about the variables:

```{r describe-variables}
var_descriptions <- fmi2::describe_variables(obs_data$variable)
var_descriptions %>% 
  DT::datatable()
```
`obs_weather_daily()` returns data  in so called  long (or melted) format 
meaning that all variable (i.e. parameter) names are  contained in column 
`variable` and corresponding values in `value` column. You can transform the 
data into a wide format using `tidyr`:

```{r spread-data}
wide_data <- obs_data %>% 
  tidyr::spread(variable, value) %>% 
  # Let's convert the sf object into a regular tibble
  sf::st_set_geometry(NULL)

wide_data %>% 
  DT::datatable()
```

Looks like there aren't too much data for `rrday`, `snow` or `TG_PT12H_min`. 
Let's have a closer look at the data:

```{r skim-data, results='asis'}
kable(skimr::skim(wide_data))
```



