---
title: "fmi2: introduction"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
vignette: >
  %\VignetteIndexEntry{fmi2 introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN", unset = "TRUE")), "true")
eval_packages <- requireNamespace(c("DT", "ggplot2", "knitr", "leaflet", "sf", "skimr"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  purl = NOT_CRAN,
  eval = NOT_CRAN && eval_packages
)
```

```{r, eval = !eval_packages, echo = FALSE}
message("Not all required packages installed. Code chunks will not be evaluated.")
```


# Introduction

This [fmi2](https://github.com/rOpenGov/fmi2) R package provides access to a subset of weather observations available from [FMI API](https://en.ilmatieteenlaitos.fi/open-data). With the package you can download data straight in R environment. The API is maintained by Finnish Meteorological Institute. The `fmi2` package is part of [rOpenGov](https://ropengov.org).

# Setup

`fmi2` is not yet in CRAN, so you'll need to install it directly from GitHub. 
While you're at it, make sure you also install all the packages below as we'll
be using them in this tutorial.

```{r installs, eval=FALSE}
install.packages(c("DT", "ggplot2", "leaflet", "remotes", "sf", "tidyverse"))

remotes::install_github("ropengov/fmi2")
remotes::install_github("ropensci/skimr")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(DT)
library(fmi2)
library(ggplot2)
library(knitr)
library(leaflet)
library(sf)
library(skimr)
```

# Getting started

You can retrieve weather station observation data with various temporal 
resolution using `fmi2`. First thing you need to know is of course which 
location exactly you want to get the data from. The FMI API provides multiple 
different ways of defining the spatial query area:

+ **Bounding box** given by coordindates and defining an area
+ **Place name** for which to provide data.
+ **FMISID** numeric FMI observation station identifier
+ **GEOID** numeric geoid of the location
+ **WMO** code of the location

We'll start off by using the **FMISID** identifies which is given to each [FMI
observation stations](https://en.ilmatieteenlaitos.fi/observation-stations). The
online table is also available in `fmi2` using the function `fmi_stations()`:

```{r show-stations}
station_data <- fmi2::fmi_stations() 

station_data %>% 
  DT::datatable()

```

We're going to pick "Hanko Tulliniemi" as an example here and use its FMISID
(100946) to retrieve the data. As you can see from the table above, it also 
provides the latlon (geographical) coordinates for the observation station.
Before we get the actual data, let's visualize Hanko region.

```{r plot-hanko-map}
# Get data for Tulliniemi only
tulliniemi_station <- station_data %>% 
  dplyr::filter(fmisid == 100946)

# Plot on a map using leaflet
leaflet::leaflet(station_data) %>% 
  leaflet::setView(lng = tulliniemi_station$lon, 
                   lat = tulliniemi_station$lat, 
                   zoom = 11) %>% 
  leaflet::addTiles() %>%  
  leaflet::addMarkers(~lon, ~lat, popup = ~name, label = ~as.character(fmisid))
```


# Getting daily weather observation data

Now that we know how the address a specific observation station, we can proceed
to getting the actual data. `fmi2` providers several functions to retrieving 
data with different variables and temporal resolution. We'll start with 
`obs_weather_daily()` which returns daily average observation data from a given 
location. Let's get the daily weather observation data for the first 6 monhts
of 2019:

```{r getting-data}
# Use Hanko Tulliniemi weather station FMISID
tulliniemi_data <- obs_weather_daily(starttime = "2019-01-01",
                                     endtime = "2019-06-30",
                                     fmisid = 100946)
```

In total, the function returned `r nrow(tulliniemi_data)` observations. You can also
note the following:

```{r sf-class}
class(tulliniemi_data)
```

which means that the data returned by `obs_weather_daily()` is a spatial `sf`
object with the `geometry` column storing the geographical information of the
weather station. We'll come back to this later. Now we are interested in what 
*kind* of data did we actually get? Let's find out:

```{r}
unique(tulliniemi_data$variable)
```

So there are six variables with their corresponding values. `fmi2` provides a
helper function `describe_variables()` that can be useful in finding out more
about the variables:

```{r describe-variables}
var_descriptions <- fmi2::describe_variables(tulliniemi_data$variable)
var_descriptions %>% 
  DT::datatable()
```
`obs_weather_daily()` returns data  in so called  long (or melted) format 
meaning that all variable (i.e. parameter) names are  contained in column 
`variable` and corresponding values in `value` column. You can transform the 
data into a wide format using `tidyr`:

```{r spread-data}
wide_data <- tulliniemi_data %>% 
  tidyr::spread(variable, value) %>% 
  # Let's convert the sf object into a regular tibble
  sf::st_set_geometry(NULL)

wide_data %>% 
  DT::datatable()
```

Looks like there aren't too much data for `rrday`, `snow` or `TG_PT12H_min`. 
Let's have a closer look at the data:

```{r skim-data, results='asis'}
(skimr::skim(wide_data))
```

Seems like the above mentioned variables indeed don't have data between the
defined days. Let's get the same data from a couple of other observation 
stations around finland. Note that this time we're using place name instead of
a FMISID.

```{r multiple stations, warning=FALSE}
oulu_data <- obs_weather_daily(starttime = "2019-01-01",
                               endtime = "2019-06-30",
                               place = "Oulu")
nuorgam_data <- obs_weather_daily(starttime = "2019-01-01",
                                  endtime = "2019-06-30",
                                  place = "Nuorgam")
# Add location name to each data set and combine them
oulu_data$location <- "Oulu"
nuorgam_data$location <- "Nuorgam"
tulliniemi_data$location <- "Hanko"

all_data <- rbind(tulliniemi_data, oulu_data, nuorgam_data)
# Factorize location and make order explicit
all_data <- all_data %>% 
  dplyr::mutate(location = factor(location, 
                                  levels = c("Nuorgam", "Oulu", "Hanko"),
                                  ordered = TRUE))
```

Let's plot the daily temperature data in different locations:

```{r plot-data}
all_data %>% 
  dplyr::filter(variable == "tday" | variable == "tmax" | variable == "tmin") %>% 
  ggplot(aes(x = time, y = value, color = variable)) + 
  geom_line() + facet_wrap(~ location, ncol=1) + ylab("Temperature (C)\n") +
  xlab("\nDate") + theme_minimal()

```

# Getting hourly weather observation data

Instead of daily values, it is also possible to retrieve weather observation
data with finer temporal resolution, such as hourly data, using the function
`obs_weather_hourly()`. The data retrieved this has slightly different content
as compared to the daily data:

```{r hourly-data}
# Get the hourly observations for the first day of 2019 in Hanko Tulliniemi
tulliniemi_data <- fmi2::obs_weather_hourly(starttime = "2019-02-01",
                                            endtime = "2019-02-02",
                                            fmisid = 100946)
```

Again, let's first have a look at what we actually got:

```{r describe-variables-2}
var_descriptions <- fmi2::describe_variables(tulliniemi_data$variable)
var_descriptions %>% 
  DT::datatable()
```
We can see that the hourly data includes more variables when compared to daily data. Hourly data for example also includes observations about wind speed. We also see that the variable names differ from the daily data. Let's compare wind speed at the same locations as before:

```{r multiple stations 2, warning = FALSE}
oulu_data <- obs_weather_hourly(starttime = "2019-02-01",
                               endtime = "2019-02-02",
                               place = "Oulu")
nuorgam_data <- obs_weather_hourly(starttime = "2019-02-01",
                                  endtime = "2019-02-02",
                                  place = "Nuorgam")
# Add location name to each data set and combine them
oulu_data$location <- "Oulu"
nuorgam_data$location <- "Nuorgam"
tulliniemi_data$location <- "Hanko"

all_data <- rbind(tulliniemi_data, oulu_data, nuorgam_data)
# Factorize location and make order explicit
all_data <- all_data %>% 
  dplyr::mutate(location = factor(location, 
                                  levels = c("Nuorgam", "Oulu", "Hanko"),
                                  ordered = TRUE))
```

Let's plot the hourly wind data in different locations:

```{r plot-data-2}
all_data %>% 
  dplyr::filter(
    variable == "WS_PT1H_AVG" | variable == "WS_PT1H_MAX" | variable == "WS_PT1H_MIN") %>% 
  ggplot(aes(x = time, y = value, color = variable)) + 
  geom_line() + facet_wrap(~ location, ncol=1) + ylab("Wind speed (m/s)\n") +
  xlab("\nDate") + theme_minimal()

```

# Getting monthly weather observation data

In addition to hourly and daily weather observations we can get monthly weather observations using function `obs_weather_monthly()`. The data obtained by the function contains a lot less variables compared to the other two functions.

```{r monthly-data}
# Get the monthly observations for year 2019 in Hanko Tulliniemi
tulliniemi_data <- fmi2::obs_weather_monthly(starttime = "2019-01-01",
                                            endtime = "2019-12-31",
                                            fmisid = 100946)
```

Let's look at what we got:

```{r describe-variables-3}
var_descriptions <- fmi2::describe_variables(tulliniemi_data$variable)
var_descriptions %>% 
  DT::datatable()
```
As we can see from the table the data only has two variables. Let's plot monthly precipitation in Turku, Vaasa and Helsinki during 2019.

```{r multiple stations 3, warning = FALSE}
turku_data <- obs_weather_monthly(starttime = "2019-01-01",
                               endtime = "2019-12-31",
                               place = "Turku")
vaasa_data <- obs_weather_monthly(starttime = "2019-01-01",
                                  endtime = "2019-12-31",
                                  place = "Vaasa")
helsinki_data <- obs_weather_monthly(starttime = "2019-01-01",
                                  endtime = "2019-12-31",
                                  place = "Helsinki")
# Add location name to each data set and combine them
turku_data$location <- "Turku"
vaasa_data$location <- "Vaasa"
helsinki_data$location <- "Helsinki"

all_data <- rbind(turku_data, vaasa_data, helsinki_data)
# Factorize location and make order explicit
all_data <- all_data %>% 
  dplyr::mutate(location = factor(location, 
                                  levels = c("Turku", "Vaasa", "Helsinki"),
                                  ordered = TRUE))
```

Let's plot the monthly precipitation data in different locations:

```{r plot-data-3}
all_data %>% 
  dplyr::filter(
    variable == "rrmon") %>% 
  ggplot(aes(x = time, y = value)) + 
  geom_line(col = "lightblue") + facet_wrap(~ location, ncol=1) +
  ylab("Precipitation amount (mm)\n") +
  xlab("\nDate") + theme_minimal()

```

# Noteworthy things

There are few things you should know when using the package. One is that there is a limit for the size of data FMI API will return on a single query. For hourly observations maximum is 744 hours (31 days), for daily observations maximum is 8928 hours (372 days) and for monthly observations maximum is 87600 hours (3650 days). So if you want data from longer time span you have to do multiple function calls.

Another good thing to know that not all weather stations will be able to provide all types of observations. This is usually apparent when a function returns a warning about there being no layers in the datasource. The warning looks like following:

```{}
> obs_weather_daily(fmisid = 103097)
Error: No layers in datasource.
```

In these scenarios it is advised to try another weather station for the wanted data. Also the data can occasionally include lot of missing values.

When using `place` as an argument for location, it is good to acknowledge that you can't pick a specific weather station from the location. So if there are several weather stations at the location, `place` argument can give data from only one of them. This can a problematic especially in the scenario the station given by `place` argument doesn't have data for wanted observations. Because of this, it is advised to use `fmisid` for location over `place`.

# Licensing and Citations

### FMI data

Cite FMI and link to [https://en.ilmatieteenlaitos.fi](https://en.ilmatieteenlaitos.fi).

* [Full license and terms of use](https://en.ilmatieteenlaitos.fi/open-data).

Central points:

* FMI open data is licensed under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
* FMI is owner of the data, unless otherwise stated contractually.

### fmi2 R package

If you use data obtained through fmi2 package in your publication, we wish that you cite the package as follows:

```{r citation, message=FALSE, warning=FALSE}
citation("fmi2")
```

Referring to the software and algorithms used in the research and identifying and acknowledging their relevance is important acording to second FORCE11 release, FORCE11 Software Citation Principle ([Smith et al., 2016](https://doi.org/10.7717/peerj-cs.86)).

# Package development, sugestions and bug reports

This package can be freely used, modified and distributed under the [MIT License](https://en.wikipedia.org/wiki/MIT_License). 

You can check the package [GitHub issues](https://github.com/rOpenGov/fmi2) for known bugs, shortcomings and other constraints. You can also use it to report bugs and to make suggestions for improving the package. User contributions are also very welcome.

# Session info

This vignette was created with 

```{r sessioninfo, warning = FALSE}
sessionInfo()
```

